---
title: "STADyUM Tutorial: Estimating Transcription Rates from Human CD4+ PRO-seq Data"
author: "Xin Zeng"
date: "2026/06/02"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
---

# Introduction

This tutorial walks through a complete analysis of human CD4+ T cell PRO-seq data
using the **STADyUM** package. STADyUM models RNA Polymerase II (Pol II) dynamics
by jointly estimating pause release and elongation kinetics from the read distribution
across promoter-proximal pause sites and gene bodies.

STADyUM estimates four key parameters per gene:

- **beta (β)**: the pause release rate — how quickly Pol II exits the
  promoter-proximal paused state. Higher beta means faster pause release.
- **chi (χ)**: a measure of overall transcriptional activity, proportional to
  both the initiation rate and elongation rate.
- **fkMean (μ)**: the mean elongation time across the gene body, reflecting
  average Pol II speed.
- **fkSD (σ)**: the standard deviation of elongation time, capturing variability
  in Pol II speed. Genes fall into two groups — "Sharp" (low σ, homogeneous speed)
  and "Broad" (high σ, heterogeneous speed).

In this tutorial we will:

1. Load TSS annotations and select one representative TSS per gene.
2. Build promoter-proximal pause and gene body read-counting regions.
3. Run `estimateTranscriptionRates()` to estimate β, χ, μ, and σ.
4. Post-process estimates and classify genes by quantile groups.
5. Visualize relationships among estimated parameters.
6. Relate β to +1 nucleosome occupancy using Micro-C 1D signal.
7. Relate β to promoter histone modification levels (H3K4me3, H3K27ac).

---

# Setup

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center"
)

library(STADyUM)
library(GenomicRanges)
library(plyranges)
library(dplyr)
library(tidyverse)
library(pracma)          # findpeaks() used in find_valley_threshold()
library(ggpointdensity)
library(cowplot)
library(patchwork)
library(RColorBrewer)
library(genomation)      # ScoreMatrix() for nucleosome signal extraction
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(rtracklayer)     # import.bw() for ENCODE bigWig files
library(ggpubr)          # stat_compare_means() for histone boxplots
```

## Configure Paths

All paths are derived from `PROJECT_ROOT`. Set that environment variable — or
update `root_dir` — to point to your local copy of the project. `load_config.R`
populates `.paths` with the standard sub-directories used throughout the pipeline.

```{r paths}
root_dir <- normalizePath(
  Sys.getenv("PROJECT_ROOT", path.expand("~/Desktop/project/YiXin_Likelihood")),
  mustWork = FALSE)
source(file.path(root_dir, "analysis", "load_config.R"))

# Helper functions shared across tutorial sections
source(file.path(root_dir, "codes", "publish", "examples", "helper_function.R"))

# ── Required input files ───────────────────────────────────────────────────────
# TSN: GRanges of detected transcription start sites with PRO-seq signal scores
tsn_file <- file.path(.paths$data, "proseq/tidgrng/PROseq-HUMAN-CD4_tsn.RDS")

# TQ: transcript annotation used to derive 3'-end (TTS) positions
tq_file  <- file.path(.paths$data, "proseq/tq/HUMAN.RDS")

# Stranded PRO-seq bigWig files (5'-end read density at nucleotide resolution)
bw_plus  <- file.path(.paths$data, "proseq/bigwig/p3/PROseq-HUMAN-CD4_plus.bw")
bw_minus <- file.path(.paths$data, "proseq/bigwig/p3/PROseq-HUMAN-CD4_minus.bw")

# ── Optional external files ────────────────────────────────────────────────────
# Micro-C 1D contact signal for +1 nucleosome analysis
ns_signal_file <- file.path(.paths$microc_1d, "human_cd4_1d_signal.bed")

# ENCODE histone modification metadata (produced by the singleCellType pipeline)
histone_meta_file <- file.path(.paths$data, "encode/meta_fc.RDS")

# Flags used to skip optional sections when external files are absent
run_nucleosome <- file.exists(ns_signal_file)
run_histone    <- file.exists(histone_meta_file)

# ── Output directory ───────────────────────────────────────────────────────────
# Only the main STADyUM rate object is saved; intermediate objects are not.
rate_dir <- file.path(.paths$outputs,
                      "publish/singleCellType/1_transcriptionRate/human")
dir.create(rate_dir, recursive = TRUE, showWarnings = FALSE)
```

---

# Load and Prepare TSS Annotations

## Load the TSN File

The TSN file is a `GRanges` object of detected transcription start sites. Each
range is linked to a gene via `ensembl_gene_id`. The `score` column reflects the
PRO-seq signal at that position; only sites with score ≥ 5 will be used in
downstream region construction.

```{r load-tsn}
cd4_tsn <- readRDS(tsn_file)
head(as.data.frame(cd4_tsn)[, c("seqnames", "start", "end", "strand",
                                 "ensembl_gene_id", "score")], 5)
```

## Select the Most Upstream TSS per Gene

Genes can have multiple annotated TSSs from alternative promoters. We keep only
the most upstream TSS per gene to provide a single, unambiguous anchor for each
locus. "Most upstream" is defined as the smallest coordinate on the `+` strand
and the largest on the `−` strand.

```{r keep-upstream-tss}
cd4_tsn <- keep_upstream_tss(cd4_tsn)
cat("TSS sites retained:", length(cd4_tsn), "\n")
```

---

# Build Read-Count Regions

STADyUM requires two counting windows per gene:

- **Pause region**: a 200 bp window starting at the TSS, capturing
  promoter-proximal paused Pol II.
- **Gene body region**: the span from the TSS to the transcription termination
  site (TTS), trimmed by 2000 bp at each end to exclude the pause peak and
  3'-processing signal. Only genes with a trimmed body ≥ 6000 bp are kept.

TSN sites with a PRO-seq score below 5 are excluded as unreliable anchors.

```{r build-regions}
tq     <- readRDS(tq_file)
region <- build_readcount_regions(cd4_tsn, tq@transcripts)

cat("Pause regions:     ", length(region$pause),    "\n")
cat("Gene body regions: ", length(region$gene_body), "\n")
```

---

# Estimate Transcription Rates

`estimateTranscriptionRates()` counts PRO-seq reads in each pause and gene body
window and fits the STADyUM kinetic model. The function returns a `STADyUM`
object containing per-gene estimates of β, χ, μ, and σ.

```{r run-stadyum}
cd4_rate <- estimateTranscriptionRates(
  bw_plus, bw_minus,
  region$pause, region$gene_body,
  "CD4"
)
cat("Genes with rate estimates:", nrow(rates(cd4_rate)), "\n")
```

---

# Post-Process Estimated Rates

We add three pieces of metadata to the rates table:

- **betaGroup / fkMeanGroup / chiGroup**: quintile (Q1–Q5) or tertile
  (Low / Medium / High) labels for each parameter.
- **fkSD**: elongation time standard deviation, computed as
  $\sigma = \sqrt{\text{fkVar}}$.
- **sdGroup**: Sharp / Broad classification. The fkSD histogram is bimodal;
  `process_sample_rate()` finds the valley between the two peaks and uses it
  as the class boundary.

```{r post-process}
cd4_rate@rates <- process_sample_rate(rates(cd4_rate))
```

## Save STADyUM Output

The estimated rate object is saved as an RDS file. It can be loaded in
downstream analyses (between-cell-type comparisons, species comparisons, etc.)
without re-running the estimation step.

```{r save-rates}
saveRDS(cd4_rate, file.path(rate_dir, "cd4_rate.RDS"))
cat("Saved:", file.path(rate_dir, "cd4_rate.RDS"), "\n")
```

The saved object contains:

- `rates(cd4_rate)`: per-gene data frame with β, χ, μ, σ, and group labels.
- `@pauseRegions`: the 200 bp counting windows used.
- `@counts`: raw pause and gene body read counts.

---

# Visualize Parameter Relationships

```{r extract-rates}
cd4_rate_df <- rates(cd4_rate)
```

## β vs χ

Each point is one gene. β and χ are positively correlated: genes with faster
pause release tend to have higher overall transcriptional output, consistent
with coupled regulation of promoter pausing and transcription initiation.

```{r beta-chi, fig.width=4, fig.height=3.5}
plot_beta_vs_chi(
  cd4_rate_df,
  label_x = -3.5, label_y = -3.8,
  xlim    = c(-6, -1),
  ylim    = c(-4, 1)
)
```

## β vs σ

The 2D density plot reveals two gene populations. The horizontal dashed line
marks the Sharp / Broad boundary identified from the bimodal fkSD histogram.
Sharp genes (below the line) have homogeneous Pol II elongation speed; Broad
genes (above the line) have more variable elongation.

```{r beta-sigma, fig.width=4.5, fig.height=4,dev='png'}
#plot_beta_vs_sigma_density(cd4_rate_df)
```

## μ vs σ

Genes with longer mean elongation times (high μ) tend to have greater
elongation-speed variability (high σ), further reflecting the Sharp / Broad
structure.

```{r mu-sigma, fig.width=4, fig.height=3.5}
plot_fkMean_vs_fkSD(cd4_rate_df)
```

---

# +1 Nucleosome Occupancy

The +1 nucleosome sits just downstream of the TSS and influences Pol II pausing.
We use Micro-C 1D contact signal as a proxy for nucleosome occupancy and ask
whether β predicts +1 nucleosome positioning.

We center a 2000 bp window around each TSS, extract the Micro-C signal from the
1D contact file, and apply a LOESS smooth (span = 0.05) to each row to reduce
positional noise while preserving the +1 peak shape.

```{r nucleosome-skip, echo=FALSE, eval=!run_nucleosome}
cat("Micro-C 1D signal file was not found,",
    "so the nucleosome section is skipped in this demo.\n")
```

```{r build-nucleosome-matrix, eval=run_nucleosome}
# Human seqinfo required by ScoreMatrix for chromosome-level range validation
seqlevelsStyle(TxDb.Hsapiens.UCSC.hg38.knownGene) <- "Ensembl"
seqinfo_human <- seqinfo(TxDb.Hsapiens.UCSC.hg38.knownGene)

ns_matrix <- get_matrix_from_nucleosome_signal(cd4_rate, ns_signal_file, seqinfo_human)

# Attach rate metadata (betaGroup, sdGroup, chiGroup) to the matrix object
ns_matrix$rate_df <- cd4_rate@rates

# Relabel chi tertiles for display
ns_matrix$rate_df$chiGroup <- factor(
  ns_matrix$rate_df$chiGroup,
  levels = levels(ns_matrix$rate_df$chiGroup),
  labels = c("Low expression", "Medium expression", "High expression")
)
cat("Nucleosome matrix dimensions:", dim(ns_matrix$smooth), "\n")
```

## β groups stratified by expression level

Each panel corresponds to one expression (χ) tertile; lines are colored by β
quintile (Q1 = slowest, Q5 = fastest pause release). Higher-β genes tend to
show a sharper +1 nucleosome peak, consistent with faster pause release being
associated with more organized chromatin downstream of the TSS.

```{r plot-beta-nucleosome, eval=run_nucleosome, fig.width=8, fig.height=3}
make_microc_profile_plot(ns_matrix, group_col = "betaGroup", class_col = "chiGroup")
```

## σ groups (Sharp vs Broad) stratified by expression level

Green = Sharp (low σ), purple = Broad (high σ). Sharp genes consistently show
a higher and more localized +1 nucleosome peak across all expression levels,
suggesting that homogeneous Pol II elongation speed correlates with well-positioned
nucleosomes.

```{r plot-sigma-nucleosome, eval=run_nucleosome, fig.width=8, fig.height=3}
sd_color <- c("#1b9e77", "#762a83")

make_microc_profile_plot(
  ns_matrix,
  group_col    = "sdGroup",
  class_col    = "chiGroup",
  colors       = sd_color,
  legend_title = expression(sigma ~ group)
)
```

---

# Histone Modifications

H3K4me3 marks active promoters and H3K27ac marks active regulatory elements.
We use ENCODE ChIP-seq fold-change bigWig files for CD4+ T cells to ask whether
β is associated with promoter histone modification levels, above and beyond what
is explained by overall expression (χ).

For each gene we define a 500 bp window centered on its TSS and compute the
mean ChIP-seq signal within that window.

```{r histone-skip, echo=FALSE, eval=!run_histone}
cat("Histone metadata file was not found,",
    "so the histone modification section is skipped in this demo.\n")
```

```{r histone-analysis, eval=run_histone}
HISTONE_MARKS <- c("H3K4me3", "H3K27ac")
CHI_LABELS    <- c("Low expression", "Medium expression", "High expression")

# Load ENCODE metadata and update bigWig paths to the local data directory
meta_fc     <- readRDS(histone_meta_file)
meta_fc$bw  <- file.path(.paths$encode_bw, basename(meta_fc$bw))
meta_cd4    <- meta_fc %>%
  filter(Biosample == "CD4 regulatory", Target %in% HISTONE_MARKS)

# Build 500 bp promoter windows and overlap with H3K4me3 / H3K27ac bigWigs
promoter        <- get_promoter_df_from_stadyum_obj(cd4_rate, promoter_len = 500)
promoter$df     <- get_histone_modification_sum(promoter$df, promoter$grang, meta_cd4)
levels(promoter$df$chiGroup) <- CHI_LABELS
```

## β vs H3K4me3

Within each expression tertile, genes with higher β carry significantly more
H3K4me3 at their promoters, linking faster pause release to a more active
promoter chromatin state.

```{r plot-h3k4me3, eval=run_histone, fig.width=8, fig.height=2.5}
plot_histone_boxplot(promoter$df, "H3K4me3", ylim = c(2, 9), label_x = 1.5)
```

## β vs H3K27ac

H3K27ac shows a similar trend: higher-β genes have more H3K27ac signal at their
promoters across all expression levels, further supporting the association
between fast pause release and an open, active chromatin environment.

```{r plot-h3k27ac, eval=run_histone, fig.width=8, fig.height=2.5}
plot_histone_boxplot(promoter$df, "H3K27ac", ylim = c(-0.5, 6.5), label_x = 1.5)
```

---

# Summary

This tutorial demonstrated a complete STADyUM workflow for human CD4+ T cells,
from raw PRO-seq bigWig files to estimated transcription kinetics and their
chromatin context. The key steps were:

1. **TSS selection** — one representative (most upstream) TSS per gene.
2. **Region construction** — 200 bp pause windows and trimmed gene body regions.
3. **Rate estimation** — `estimateTranscriptionRates()` returning β, χ, μ, σ.
4. **Post-processing** — quintile groups and Sharp / Broad classification saved
   in the `rates` data frame of the `STADyUM` object.
5. **Parameter relationships** — β–χ positive correlation, bimodal β–σ structure,
   μ–σ positive correlation.
6. **Chromatin context** — higher-β genes show stronger +1 nucleosome peaks
   (Micro-C) and higher H3K4me3 / H3K27ac (ENCODE ChIP-seq) at their promoters
   (when external data are available).

This HTML report is designed to show the analysis and its figures inline in a
single self-contained file. The full manuscript pipeline (scripts in
`singleCellType/`) runs the same analyses for all species and cell types and
saves publication-quality PDF figures separately.

**Downstream analyses** enabled by the saved `cd4_rate.RDS`:

- **Between-cell-type comparison**: pair CD4 and CD14 rate objects and apply a
  likelihood ratio test to identify genes with cell-type-specific changes in β
  or χ.
- **Between-species comparison**: align orthologous TSSs across primate species
  and test for evolutionary divergence in pause release kinetics.
- **Sequence feature integration**: link β or σ groups to core promoter motifs
  (TATA box, GC box, Initiator) or GC content to understand the sequence
  determinants of Pol II pausing.

---

# Session Information

```{r session-info}
sessionInfo()
```
